---
title: "The Blunder Radar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Blunder Radar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

An engine answers "what is best here". That is not the question you have when
you are about to move. The useful question is "what am I likely to get wrong",
and the answer depends on who is playing.

## The idea

Two things are already available: how likely a human of a given strength is to
play each move, and what each move costs. Multiply them.

```
risk = SUM over moves of  P(a human plays it) x centipawns it loses
```

The first term comes from [Maia](https://github.com/CSSLab/maia-chess), a
network trained to predict human moves at a target rating rather than to win.
The second comes from Stockfish.

A move that hangs a queen but no human would consider carries no risk. A natural
move that quietly drops a pawn carries a lot. Neither an engine nor a
human-move model gets you that on its own.

## Why the engine has to be asked directly

The moves a human is likely to blunder into are, by definition, ones the engine
ranks poorly - so they fall outside an ordinary MultiPV list and would never be
scored. `evaluate_moves()` uses UCI `searchmoves` to make the engine evaluate
exactly the candidate moves, whatever it thinks of them:

```{r}
maia <- maia_session_start(1500)
risk <- blunder_risk(
  "r1bqkbnr/pppp1ppp/2n5/4p2Q/2B1P3/8/PPPP1PPP/RNB1K1NR b KQkq - 3 3",
  maia
)
risk$risk
#> [1] 86.9
head(risk$moves)
```

## Does it work?

A risk metric that fires everywhere says nothing, so it was measured against
known traps *and* a quiet control:

| Position | Risk @1100 | Risk @1900 |
| --- | ---: | ---: |
| Legal's mate | 6724 cp | 9192 cp |
| Scholar's mate | 1673 cp | 998 cp |
| Blackburne | 90 cp | - |
| **Starting position** | **3.0 cp** | **3.6 cp** |

Roughly 2000x between a trap and a position where nothing can go wrong. The
bottom row is the one that makes the rest meaningful.

A result worth noticing: Legal's mate scores **higher** at 1900 than at 1100.
The stronger player is *more* confident about grabbing the queen - 93.3% versus
65.7% - so the trap works better on them. That is not something an engine
evaluation can express.

## The arrows

On the board the engine's choice is blue and the human moves are orange, or red
where they lose real material. Which moves get an arrow is chosen by **risk
contribution**, not by raw probability.

That distinction is the whole feature. After `3.Qh5`, Black's losing `Nf6` is
only the *fourth* most probable reply. Drawing the three most likely moves
produces three orange arrows and quietly omits the one move worth warning
about. `radar_arrow_moves()` takes the top contributors and adds the single
likeliest move for context.

## Trappiness

The same machinery, pointed the other way: for each move *you* could play, how
badly is the opponent likely to go wrong afterwards?

```{r}
trappiness(fen, maia, ctx)
```

`practical` is what the move is likely to win you from their mistakes, minus
what it costs you objectively. A slightly inferior move that sets a problem
they will probably fail can be the better practical choice.

## Requirements

The radar needs lc0 and the Maia weights. On Windows `ensure_maia()` fetches
both. There is no official Linux build of lc0, so the container image compiles
it from source - see the `Dockerfile`. Without them the app still runs and the
radar reports that the human model is unavailable rather than failing.
