---
title: "Following a live game"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Following a live game}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Analysing one frozen position is useful. Following a game as it happens is more
useful, and much harder, because the input is a stream of imperfect readings of
a board that is sometimes mid-animation.

## The one rule

A position joins the game only when **exactly one sequence of legal moves
explains how it got there**.

That single gate replaces a pile of filters:

- a misread square produces a position no legal move reaches, so it is rejected;
- a frame caught mid-animation is not a legal successor either, so it is
  rejected;
- a ply that was never captured on screen is recovered, because a two-move
  sequence explains the jump;
- castling, en passant and promotion are handled correctly for free, because
  chess.js generates the moves rather than a square-by-square diff guessing at
  what changed.

```{r}
ctx <- new_chess_context()
game <- game_new()
res <- track_observation(ctx, game, observed_fen, max_plies = 2)
res$status
#> [1] "accepted"
game <- game_accept(ctx, game, res)
```

The ambiguity requirement matters as much as the legality one. If *two*
different move sequences could explain a new position, the reading is not used
- guessing which one happened would silently corrupt the record, and a tracker
that invents moves is worse than one that admits it lost the thread. Readings
that cannot be explained are counted and shown, not quietly dropped.

## Orientation and turn, settled by the game

A screenshot cannot show whose move it is, and can be ambiguous about which way
round the board is. Rather than guess from pixels, both candidate readings go
through the same gate, and the one that continues the game legally is the one
that was right.

This is why the live tracker needs no "side to move" control while a game is
running, when analysing a single screenshot does.

## Cost

Recognition is not free, so frames are compared in the browser and only a
settled, changed picture is sent to the server - roughly one recognition per
move rather than one per tick.

Searching deeper for missing plies costs more than it sounds: a 2-ply search
takes about 0.12 s, a 3-ply search about 4.8 s. The UI caps at 2 for that
reason.

## What a browser can actually do

Capture uses `getDisplayMedia`: the browser's own dialog, one explicit choice of
screen, window or tab, and the board marked once by dragging over a still frame.

Mark the region roughly on the board's edge. It is grown outward slightly before
use, because slicing *into* the board is fatal to recognition - cutting 2 px in
takes match quality from 0.94 to 0.17. A very loose region fails differently:
past a margin of about 5% the edge detection stops finding the board at all.

Two things a web page genuinely cannot do, and which are not pretended at here:

- **read the screen unprompted** - there is always a permission dialog;
- **watch the clipboard in the background** - `navigator.clipboard.read()` is
  refused unless the document has focus, which it will not have while you are
  playing in another window.

So pasting a screenshot after each move is offered as a first-class mode rather
than a fallback. It goes through the identical path.

One practical constraint: the app's window has to stay visible. Browsers
throttle timers in hidden tabs, which stalls the capture loop.

## Reviewing the game

Once there are moves, three views open up:

- an **evaluation graph** across the whole game;
- **where the game turned** - the moves that cost the most;
- clicking any move rewinds the board to just before it, and offers to ask the
  Blunder Radar whether that mistake was predictable.

That last one is where this and the radar meet. After `3.Qh5 Nf6` it reports
that a 1500 player plays `Nf6` here 6% of the time and that it is the single
largest source of risk in the position - what the move cost, and why it was
tempting, together.
