---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Tanmai Chess reads a chess board from a screenshot and tells you the best move.
It also does something engines do not: it tells you where a player of a given
strength is likely to go **wrong**.

Start it with:

```{r}
library(tanmai)
run_app()
```

Or use the deployed copy at
<https://tanmai-103985161147.us-central1.run.app/>, which already has every
engine and network built in.

## Overview

The app has two tabs. Nearly everything happens on the first.

| Tab | What it is for |
| --- | --- |
| **Analyse** | Read a screenshot, explore the position on a live board, follow a game as you play it |
| **Piece sets** | Download the Lichess piece sets, or teach the app art it has never seen |

The pipeline underneath is:

```
screenshot -> 64 squares -> recognise pieces -> FEN -> Stockfish -> best move
                                                |
                                                +-> blunder radar
                                                +-> live game tracking
```

## 1. Reading a position

Drop a screenshot onto **Read a position**, or click the paste zone and press
`Ctrl+V`. A file picker is there too.

Three settings sit beside it:

- **Side to move** - a still image cannot show whose turn it is, so it has to be
  told. (In a tracked game it is worked out for you; see below.)
- **Orientation** - leave it on *Auto* unless it gets it wrong. Auto decides
  from where the two armies sit, which is reliable for real positions.
- **Auto-crop to the board** - trims the surrounding page. Leave it on unless
  your screenshot is already cropped tight.

Press **Analyse position**. What was read appears beside the screenshot: the
piece set it matched, which way round it decided the board is, and the FEN.

If the piece set is not one it knows, it says so rather than guessing:

> **Unrecognized piece set.** This board doesn't match any installed piece set
> well... so the reading below is an unreliable guess.

That usually means a Chess.com or custom theme. Two ways forward: download the
Lichess sets (§4), or just drag the misread pieces into place on the board -
the position is loaded there either way.

## 2. Exploring on the board

The recognised position lands on the **Analysis board** automatically. There is
no button to press.

Move pieces by dragging, or by tapping: tap a piece and its legal destinations
appear as dots, captures as rings, then tap where it should go. Tapping is
usually easier on a phone, where a square is about 34 pixels wide.

Alongside the board:

- the **evaluation** and the bar down the left, updating as the search deepens;
- the engine's **top three lines**;
- the **move list** and the current **FEN**.

`Flip`, `Undo`, `Start` and `Play best move` are under the board.

## 3. The Blunder Radar

Tick **Show what a human would play** and pick an opponent rating.

An engine tells you what is *best*. The radar answers a different question: of
the moves a human of this strength might actually pick, which ones cost
something, and how likely are they? It scores every legal move by

```
P(a human plays it) x centipawns it loses
```

using [Maia](https://github.com/CSSLab/maia-chess) for the first term and
Stockfish for the second.

On the board, the engine's choice is drawn in **blue** and the human moves in
**orange** or **red**. The arrows are chosen by how much each move contributes
to the risk, not by how likely it is - in a trap the dangerous move is by
construction *not* the most likely one.

`vignette("blunder-radar")` goes into how it is scored and how it was tested.

## 4. Piece sets

Open the **Piece sets** tab and press **Download piece sets** once. That fetches
38 Lichess sets (about 1 MB) into a local cache; screenshots using any of them
are then recognised with nothing to configure.

For art the app cannot know - Chess.com themes, custom pieces - use **Teach an
unknown set**: paste a screenshot of that site's *standard starting position*
and press **Calibrate**. The result joins auto-detection as the `custom` set.

Chess.com's art is deliberately not downloaded: it is proprietary and its terms
prohibit scraping. Manual calibration exists for exactly this.

## 5. Following a live game

The **Follow a live game** panel tracks a game as it is played.

Press **Share a screen**, choose the window you are playing in, then drag a
rectangle over the board on the still frame that appears. Mark it roughly on the
board's edge - a little outside is fine and better than slicing into it.

Or skip sharing entirely and paste a fresh screenshot after each move. Both feed
the same tracker.

As the game goes on you get the move list, an evaluation for each move, a graph
across the whole game, and a list of **where the game turned**. Clicking any
move rewinds the board to just before it and offers to ask the radar whether
that mistake was predictable.

A position is only added when exactly one sequence of legal moves explains it,
so misreads and half-drawn animation frames are discarded rather than corrupting
the record. `vignette("live-tracking")` explains why that one rule does so much
work.

## 6. Estimating the opponent

When reviewing a move, the strength selector offers **Estimated from this
game**. Rather than asking you how strong your opponent is, it infers it: each
Maia network is a distribution over moves, so the network least surprised by
what they actually played is the most likely one.

It is right about 75% of the time on its own, which is better than guessing
between three networks but is not a fact. So it declines to answer unless the
moves genuinely discriminate - then it is right about 94% of the time, and says
so. When it cannot tell, it says that too and falls back to 1500.

## What it cannot do

- **Whose turn it is** cannot be read from a still image.
- **En passant** is not inferred, and **castling rights** are assumed only when
  king and rook are on their home squares.
- A web page cannot read your screen unprompted or watch the clipboard in the
  background. Screen capture uses the browser's own permission dialog, and the
  app's window has to stay visible - browsers throttle timers in hidden tabs.
- A few Lichess sets are genuinely hard to tell apart because their *own* pieces
  are nearly identical; `mono`, `letter` and `disguised` are excluded outright.
